Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-02 analysis/paper/article.qmd
updating section on background to the wiki

library(tidyverse)
library(here)
library(ggbeeswarm)
library(flextable)
# This CSV file was downloaded from our data sheet here
# https://docs.google.com/spreadsheets/d/1Jwe3UqJyedrV-QWlwR_44__t4xBVrCfxGyhXdi3E0sg/edit?resourcekey#gid=1686084773
# note that you may need to download it again to get the latest updates!

@@

# scrape the web pages to get the 
# ratio of tenure-track to untenured positions
# base URL changes after 2018_2019
# We did this once and then saved the output so we don't scrape every
# time the quarto document is rendered. So this code block is
# eval: false to avoid repeated and unnesecary scraping

@@

@@

```{r}
#| label: tbl-show-basic-counts
#| tbl-cap: "Breakdown of counts of job ads by rank and tenure status. Values in parentheses are for the United States only. TT = tenure-track, NTT = non-tenure-track. Not all ads include unambiguous information about tenure status, so the sum of TT and NTT does not equal the sum of all jobs in all ranks."

>>>>

library(tidyverse)
library(here)
library(ggbeeswarm)
library(flextable)
library(rvest)
# This CSV file was downloaded from our data sheet here
# https://docs.google.com/spreadsheets/d/1Jwe3UqJyedrV-QWlwR_44__t4xBVrCfxGyhXdi3E0sg/edit?resourcekey#gid=1686084773
# note that you may need to download it again to get the latest updates!

@@

# scrape the web pages to get the 
# ratio of tenure-track to untenured positions
# base URL changes after 2018_2019
# We did this once and then saved the output so we don't scrape every
# time the quarto document is rendered. So this code block is
# eval: false to avoid repeated and unnecessary scraping

@@

#------------------------------------------------------
# scrape wiki page to describe editing dyanmics,
# save the results so we don't do this often

# starting from : ratio-tt-to-non-tt above
# get the history of edits page "?action=history&offset=&limit=10000"

edits_page <- "?action=history&offset=&limit=10000"

edits_urls_for_each_year <-
  str_glue('{urls_for_each_year}',
           '{edits_page}' )

# scrape from the webpages and make a table of key edit variables
# this takes a few minutes
edits_urls_for_each_year_lst_tbl <-
map(edits_urls_for_each_year,
    ~{
      # get webpage
      pge <- read_html(.x)

      # extract key variables for each edit
      # into a table
      tibble(
      edit_date =  pge %>%
        html_elements(".mw-changeslist-date") %>%
        html_text2(),
      edit_name =  pge %>%
        html_elements(".mw-userlink bdi") %>%
        html_text2(),
      edit_size =  pge %>%
        html_elements(".mw-plusminus-neg , .mw-plusminus-null , .mw-plusminus-pos") %>%
        html_text2()
      )
    }
)

# handle getting the edit comment which tells us which section
# of the page was edited, it's blank for many edits, but there
# might be some interesting patterns in there
edits_urls_for_each_year_lst_section <-
  map(edits_urls_for_each_year,
      ~read_html(.x) %>%
        html_elements('#pagehistory li') %>%
        html_text2()
  )

edits_urls_for_each_year_lst_section_tbl <-
map(edits_urls_for_each_year_lst_section,
    ~{
      # Extract time and date
      edit_date <-
        stringr::str_extract(.x, "\\d{2}:\\d{2}, \\d{1,2} [A-Za-z]+ \\d{4}")

      edit_size <- str_extract(.x, "bytes\\s([+-]?\\d+)") |>
         str_remove("bytes\\s")

      # Extract text between → and undo
      edit_comment <- stringr::str_extract(.x, "→(.*?) undo")
      edit_comment <- stringr::str_remove_all(edit_comment, "→| undo")  # clean up

      # Combine into a data frame
      tibble(edit_date,
             edit_size,
             edit_comment)
    }
)

# join together key variables and edit comments
edits_urls_for_each_year_lst_tbl_with_comments <-
map2(edits_urls_for_each_year_lst_tbl,
     edits_urls_for_each_year_lst_section_tbl,
    ~ left_join(.x, .y,
                relationship = "many-to-many") %>%
      # deduplicate
      distinct()
)

names(edits_urls_for_each_year_lst_tbl_with_comments) <-
  c(years_to_2019,
    years_after_2020)

# combine list into one big data frame
edits_for_each_year_tbl <-
bind_rows(edits_urls_for_each_year_lst_tbl_with_comments,
          .id = "year_ad_posted") %>%
  mutate(
    edit_size_num = parse_number(str_replace(edit_size, "−", "-")),
    edit_time = parse_date_time(edit_date,
                                 orders = "HM, d B Y"),
    # round down to first day of month
    edit_month = floor_date(edit_time,
                            unit = "month"),
    edit_year = floor_date(edit_month,
                           unit = "year"),
    edit_year_only = year(edit_month),
    job_market_year = parse_number(str_sub(year_ad_posted,
                              -4L))
  ) %>%
  # delete edits that were made after the job market year
  filter(edit_year_only == job_market_year | edit_year == (job_market_year -1))

write_csv(edits_for_each_year_tbl,
          here::here("analysis/data/raw_data/edits_for_each_year_tbl.csv"))

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for  three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 Bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year."

edits_for_each_year_tbl <- read_csv(
          here::here("analysis/data/raw_data/edits_for_each_year_tbl.csv"))

# visualisations ----------------------------------

# Number of edits per year
p_edits_per_year <-
ggplot(edits_for_each_year_tbl) +
  aes(as_date(edit_year)) +
  geom_bar() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# Number of unique editors per year
p_editors_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(n_distinct_editors = n_distinct(edit_name)) %>%
ggplot() +
  aes(as_date(edit_year),
      n_distinct_editors) +
  geom_col() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of editors\n(distinct usernames or IP addresses)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# Editor activity over time

# life span of an editor, from first edit to last edit
edits_for_each_year_tbl_span <-
edits_for_each_year_tbl %>%
  group_by(edit_name) %>%
  summarise(first_edit = min(edit_month),
            last_edit =  max(edit_month),
            edit_span = last_edit - first_edit,
            edit_span_years = as.numeric(edit_span,
                                         units = "days") / (365.25)) %>%
  mutate(first_edit_year = year(first_edit),
         last_edit_year =  year(last_edit))

# Most editors are only active for less that one year
p_editor_life_distr <-
ggplot(edits_for_each_year_tbl_span) +
  aes(edit_span_years) +
  geom_histogram() +
  theme_minimal() +
  scale_y_log10() +
  labs(x = "Years editing", y = "Number of editors")

# how many active for less than one year?
edits_for_each_year_tbl_span_tbl <- 
edits_for_each_year_tbl_span %>% 
  group_by(edit_span_years) %>% 
  summarise(n = n()) %>% 
  mutate(perc = n / sum(n) * 100)

edits_for_each_year_tbl_span_tbl_less_than_one_year <-
edits_for_each_year_tbl_span_tbl %>% 
  mutate(sum_less_than_one = if_else(cumsum(edit_span_years) < 1, n, 0)) %>% pull(sum_less_than_one) %>% sum()

edits_for_each_year_tbl_span_tbl_less_than_one_year_perc <- 
  round(
    edits_for_each_year_tbl_span_tbl_less_than_one_year /  sum(edits_for_each_year_tbl_span_tbl$n) * 100, 0)

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
ggplot() +
  aes(as_date(first_edit_year),
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 1,
               lineend = "butt") +
  theme_minimal() +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank())

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# which sections get the most edits?
section_edit_tally_tbl <-
edits_for_each_year_tbl %>%
  mutate(edit_comment = str_squish(edit_comment)) %>%
  filter(!is.na(edit_comment)) %>%
  # remove all caps text at the end of the position name, those are status changes
  # in the same position
  mutate(edit_comment = gsub("\\s+[A-Z]+(?:\\s+[A-Z]+)*$", "", edit_comment)) %>%
  filter(!edit_comment %in% c("Current Users",
                              "DISCUSSION, RUMORS, SPECULATION",
                              "DISCUSSION, RUMORS AND SPECULATION",
                              "General Discussion, Rumors, and Speculation",
                              "Rejection Etiquette",
                              "Current Users Edit",
                              "Description",
                              "RESEARCH",
                              "TENURE",
                              "NON-TENURE-TRACK",
                              "ASSISTANT",
                              "TENURE-TRACK OR TENURED / FULL-TIME",
                              "Tenure-Track or Tenured / Full-time Position",
                              "Posting Names of Job-Getters",
                              "VISITING POSITIONS / LIMITED TERM APPOINTMENTS / POSTDOCS")) %>%
  filter(!str_detect(edit_comment, "Updated:|Wiki")) %>%
  group_by(edit_year,
           edit_comment) %>%
  tally(sort = TRUE)

# edits per section for each year
library(ggbeeswarm)
p_edit_number_per_entry_per_year <-
ggplot(section_edit_tally_tbl) +
  aes(as_date(edit_year),
      group = as_date(edit_year),
      n) +
  geom_boxplot() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits per job ad") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

library(cowplot)
plot_grid(
  plot_grid(p_edits_per_year,
          p_editors_per_year,
          p_median_edit_size_per_year,
          p_edit_number_per_entry_per_year,
          labels = c("A", "B", "D", "E")),
          plot_grid(p_editor_life_distr_year,
                    p_editor_life_distr,
                    rel_widths = c(2),
                    ncol = 1,
                    labels = c("C", "F"))
          )
```

Because there is no single editorial authority, collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. We can get some insights into the dynamics of editing activity from the MediaWiki technology used by Academic Jobs Wiki that creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics shows the temporal trends in editing of the Academic Jobs Wiki. The number of edits and editors roughly halved from 2013 to 2017, and while the number of edits remained relatively steady after 2017, the number of editors continued to gradually decline. The vast majority of editors are active on the site for less than a year (n = r` edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads.

```{r}
#| label: fig-panel-wiki-self-report-numbers
#| fig-cap: "Number of users self-reporting in each category of 'current users' per year"

# get the dat
# 2021-22 and 2022-2023
current_user_tbl <-
map(
urls_for_each_year[1:9],
~.x %>%
  read_html() %>%
  html_elements("p+ table td") %>%
  html_text2()
)

current_user_tbl_2021_22 <-
    urls_for_each_year[10]  %>%
      read_html() %>%
      html_elements("table:nth-child(1086) td") %>%
      html_text2()

current_user_tbl_2022_2023 <-
    urls_for_each_year[11] %>%
      read_html() %>%
      html_elements("h2+ .fandom-table td") %>%
      html_text2()

current_user_tbl_all <- append(current_user_tbl,
                        c(list(current_user_tbl_2021_22),
                          list(current_user_tbl_2022_2023)
                        ))

names(current_user_tbl_all) <- urls_for_each_year

names(current_user_tbl_all)[names(current_user_tbl_all) == "https://academicjobs.fandom.com/wiki/Archaeology_2021-22"] <- "https://academicjobs.fandom.com/wiki/Archaeology_2021-2022"

wiki_data_list <-
map(current_user_tbl_all,
    ~.x %>% .[1:40] %>%
    stringi::stri_remove_empty(.x))

# Use map_dfr to iterate through the list and its names, binding results
tidy_table <- map_dfr(wiki_data_list, ~{
  # Process one vector (.x) at a time
  tibble(raw = .x) %>%
    # Create a lagged column to easily access the previous element
    mutate(
      previous_raw = lag(raw),
      # Attempt to parse the current value as a number
      # suppressWarnings handles cases where it's clearly not numeric
      # !is.na checks if parsing was successful
      is_count = !is.na(suppressWarnings(readr::parse_number(raw))),
      # Check if the *previous* value looks like a number
      prev_is_count = !is.na(suppressWarnings(readr::parse_number(previous_raw)))
    ) %>%
    # Keep rows where:
    # 1. The current row IS a count (is_count == TRUE)
    # 2. The previous row exists (previous_raw is not NA)
    # 3. The previous row IS NOT a count (prev_is_count == FALSE)
    #    This ensures we grab the status text, not a preceding number
    filter(is_count & !is.na(previous_raw) & !prev_is_count) %>%
    # Select and rename the relevant columns, converting count to numeric
    transmute(
      status = previous_raw,
      count = readr::parse_number(raw) # Convert count string to number
    )
}, .id = "url") %>% # .id creates a column 'url' from the list names
  # Extract the year from the URL
  mutate(
    year = str_extract(basename(url), "\\d{4}-\\d{4}"), # Extracts YYYY-YYYY pattern
    .before = status # Place the year column before status
  ) %>%
  select(year, status, count) # Keep only the final desired columns

#  the resulting tidy table
tidy_table_to_plot <- 
tidy_table %>%
  filter(!status %in% c("I'm a lurker",
                        "I have a current but term-limited faculty job",
                        "I have some other teaching, research, or non-academic gig",
                        "Nope (independent scholar, on sabbatical from adjuncting, etc.)")) %>%
  mutate(year = str_replace(year, "-", "-\n"))

# summary over time
how_many_people_use_the_wiki <- 
tidy_table_to_plot %>% 
filter(status == "How many people use the wiki?")

how_many_people_use_the_wiki_perc_start_end <- round(
(how_many_people_use_the_wiki$count[1] - how_many_people_use_the_wiki$count[nrow(how_many_people_use_the_wiki)]) / how_many_people_use_the_wiki$count[1] * 100, 0)

# proportion of search committee members
search_committee_members_tbl <- 
tidy_table_to_plot %>% 
  filter(status %in% c( "How many people use the wiki?",
                       "I'm a search committee member")) %>% 
  pivot_wider(names_from = status,
              values_from = count) %>% 
  mutate(perc = round(`I'm a search committee member` / 
          `How many people use the wiki?` * 100, 0)) %>% 
  drop_na()

search_committee_members_perc_start <- 
  search_committee_members_tbl$perc[1] 
search_committee_members_perc_end <- 
  search_committee_members_tbl$perc[nrow(search_committee_members_tbl)]

  ggplot(tidy_table_to_plot) +
  aes(year,
      count) +
  geom_col() +
  theme_minimal() +
  facet_wrap(~ status,
             scales = "free") +
  xlab("") +
  theme(axis.text.x = element_text(size = 6))

tidy_table_to_plot %>% 
filter(status != "How many people use the wiki?") %>% 
ggplot() +
  aes(year,
      count,
      fill = status) +
  geom_col() +
  theme_minimal() +
  xlab("") +
  scale_fill_brewer(type = "qual") +
  theme(axis.text.x = element_text(size = 6),
        legend.position = c(.7,.7))
```

There seems to be a shift over time in the balance of editors in these two groups, as indicated by the number of self-reporting users over time. @fig-panel-wiki-self-report-numbers shows the data found on each page in a section labeled 'current users', where users can volunteer to update a table tracking the status of users. These data indicate that the number of self-reporting users, from the values recording at 'How many people use the wiki?', declined by `r how_many_people_use_the_wiki_perc_start_end`% during the study period. At the same time, the proportion of users self-identifying as "search committee members" increased, from under `r search_committee_members_perc_start`% in 2014-2018 to over `r search_committee_members_perc_end`% from 2020 onward. This suggests that over time fewer job applicants have been contributing job postings and sharing information about their experiences, and search committee members were increasingly posting job advertisements themselves. Overall the Archaeology Academic Jobs Wiki is best understood not as a comprehensive archive of all jobs posted during the study period, but a biased sample of the job ads that were of most interest to archaeologists applying for jobs in a given year. It represents the jobs that early career applicants were most focused on applying for, and getting information on.

@@

```{r}
#| label: tbl-show-basic-counts
#| tbl-cap: "Breakdown of counts of job ads by rank and tenure status. Values in parentheses are for the United States only. TT = tenure-track, US-only, NTT = non-tenure-track, US and elsewhere. Not all ads include unambiguous information about tenure status, so the sum of TT and NTT does not equal the sum of all jobs in all ranks."
